---
title: "Effect of transmission type on efficiency of classic cars"
author: "Vineet W. Singh"
date: "15 April 2018"
output:
  html_document: default
  keep_md: true
  pdf_document: 
    fig_width: 6.4
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Summary**  

The mtcars data set provides technical data and efficiency for a set of 
classic cars sold in the US, prior to the fuel crises of the 1970's.  

Using this data set, this analysis tries to address the following questions:  
_1) Does the transmission type contribute to the efficiency of classic cars?_ 
and  
_2) What is the average difference in mpg of cars with_
_different transmission types?_  

The efficiency of a car, is usually measured in the average number of miles
per gallon (_mpg_) it runs and the _mpg_ depends upon a number of 
factors.
Linear models were made and all factors in the data set were analysed in a 
systemic way, to find out which variables/factors have the 
most significant effects on the efficiency (_mpg_) of these classic cars.  

It is observed that besides the transmission type(_am_), 
_mpg_ also depends upon 
the weight (_wt_) and the horse power (_hp_) of the car and this 
relationship is approximated by the following linear equation/model:  
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_**  

**Analysis**  

```{r chunk0,echo=FALSE, results='hide'}
data(mtcars)
require('car')
mdl2<-lm(mpg~factor(am)+wt,mtcars)
```
Exploratory data analysis involved making box plots (Appendix) in which the 
efficiency (in _mpg_) was grouped by the transmission type
(_am_). From the box plots itself, 
it can be seen that manual transmission cars (*MT*/_am_ = 1) are more 
efficient 
and give higher _mpg_ than automatic transmission (*AT*/_am_ = 0) 
cars. However, there is considerable variation in the _mpg_ within each 
transmission type (_am_) and effects of other factors should also be 
analysed.  
To begin with, the simplest of models is made, and this calculates as to how 
_mpg_ varies by _am_ (*AT/MT*). Computation 1 (Appendix) 
provides the first linear model: _mpg=17.147+7.245(am)_. The coefficients of 
this simple model show that a _MT_ car 
runs 24.392 _mpg_ compared to 17.145 _mpg_ for an _AT_ car.  

Any model that calculates car _mpg_, should also take into consideration, 
that efficiency of any car is proportional to it's weight. Should weight be included 
in the model?  
In computation 2 (Appendix), a new model is made in which weight 
(_wt_) is included.  
The model residuals are tested for normalility by using the 
shapiro test. The NH for the shapiro test is that the residuals follow a normal 
distribution. Shapiro results in a p-value of .10 and the NH is accepted. 
ANOVA is possible between the previous model and the new model.  
The null hypothesis (NH) proposed for ANOVA is that omitting _wt_ will not 
increase the bias in the model. Alternative hypothesis (AH) is that omitting 
_wt_ will increase the bias.  
From ANOVA, we find that the F score is 46.115 (P value <.0001), 
which is significant. We therefore, reject the 
NH, accept the AH and include _wt_ to improve the model.  
The model changes to: _mpg=37.322-0.024(am)-5.353(wt)_.  

Next we need to test, whether to include other engine variables 
(like horse power _hp_) that might effect _mpg_.  
_hp_ is added to the model and evaluated as below: 
```{r chunk1}
mdl9<-lm(mpg~factor(am)+wt+hp,mtcars)
vif(mdl2)
vif(mdl9)
shapiro.test(mdl9$residuals)
anova(mdl2,mdl9)
```
Including _hp_ increases variance of both the _wt_ and _am_ 
coefficients, so do we retain _hp_ as a regressor in our model?  
The residuals of the testing model are plotted in the Appendix and they show an 
even dispersion.The histogram of the residuals shows an approximately 
normal distribution.  
From the shapiro test, residuals of the updated model are normally distributed 
(p-value: .11) so ANOVA can be performed between this model and the previous 
one. ANOVA tests the NH that omitting the _hp_ will not increase 
bias in the model.  
ANOVA results in a F score of 15.224 (P value < .001), which is 
significant. The NH is rejected and _hp_ is included in 
the model which now transforms to: 
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_**  
From the summary  of the model (computation 3, Appendix), we can observe that 
the residual standard error is 2.538 _mpg_. 

In the final test, computation 4 (Appendix), we include all variables 
(in addition to _am , wt and hp_) in the testing model.  
The shapiro test on residuals gives a P-value of .23 which implies that the 
residuals are normally distributed and we can perform ANOVA on the testing 
model and the previous model.  
The NH for the ANOVA is: Omitting all regressors 
except _am , wt and hp_ will not increase bias in the model. 
Subsequent ANOVA between the previous model and the final testing model 
gives an F score  of 0.667 (P-value: .70) which is not significant, so we cannot
reject the NH.  
It is inferred that including other variables in the testing model 
increases the variances of the coefficients that matter but does not 
significantly reduce the RSS(180 vs 147) between the models. Therefore 
there is no gain in including any regressors other than 
_am , wt and hp_ in the model.  

**Conclusion**  

Based on the analysis, the linear model specified by 
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_** with a standard error of 
2.538 _mpg_ is the best fit to the data provided.  
We can conclude that a *MT* car will run 2.084 miles more than an *AT* car with 
the same _wt/hp_. Each ton increase in _wt_ will decrease the 
car _mpg_ by 2.879 _mpg_ and one _hp_ increase in engine power 
will decrease the car _mpg_ by 0.038 mpg.  
Approximately 66% of cars (data points) of this sample set will have an _mpg_ 
that is within an 
error margin of +/- 2.538 _mpg_ of the _mpg_ predicted by the equation/model. 

## Appendix

**EDA-Box Plot**
```{r  chunk2}
par(cex=0.7)
boxplot(mpg~factor(am),data=mtcars, main="Car Mileage Data",
    xlab="Auto (0/AT) or Manual (1/MT) Transmission ", ylab="Miles Per Gallon") 
```
  
  **Residual Plot and Residual Histogram for the final model** 
```{r chunk3}
par(mfrow=c(1,2), cex=0.5)
plot(mdl9$residuals,main="Residual Plot",xlab="Sample No.", ylab="Residual" )
hist(mdl9$residuals, main="Histogram of Residuals",ylab="Frequency",
     xlab="Residual Value")
```
**Computation 1**
```{r chunk4, eval=FALSE}
mdl1<-lm(mpg~factor(am),mtcars)
mdl1$coefficients
shapiro.test(mdl1$residuals)
```
**Computation 2**  
```{r chunk5, eval=FALSE}
mdl2<-lm(mpg~factor(am)+wt,mtcars)
mdl2$coefficients
shapiro.test(mdl2$residuals)
anova(mdl1,mdl2)
```
**Computation 3**
```{r chunk6, eval=FALSE}
summary(mdl9)
```
**Computation 4**
```{r chunk7,eval=FALSE}
mdl12<-lm(mpg~factor(am)+wt+hp+disp+cyl+drat+qsec+factor(vs)+carb+gear,mtcars)
vif(mdl12)
shapiro.test(mdl12$residuals)
anova(mdl9,mdl12)
```









 
 