---
title: "Effect of transmission type on efficiency of classic cars"
author: "Vineet W. Singh"
date: "15 April 2018"
output:
  pdf_document: 
    fig_width: 3.2
    fig_height: 2
  keep_md: true
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary
The mtcars data set provides technical data and efficiency for a set of 
classic cars sold in the US, prior to the fuel crises of the 1970's.  

Using this data set, this analysis tries to address the following questions:  
_1) Does the transmission type contribute to the efficiency of classic cars?_ 
and  
_2) What is the average difference in mpg of cars with_
_different transmission types?_  

The efficiency of a car, is usually measured in the average number of miles
per gallon (**_mpg_**) it runs and the **_mpg_** depends upon a number of 
actors.
Linear models were made and all factors in the data set were analysed in a 
systemic way, to find out which variables/factors have the 
most significant effects on the efficiency (**_mpg_**) of these classic cars.  

It is observed that besides the transmission type(**_am_**), 
**_mpg_** mainly depends upon 
the weight (**_wt_**) and the horse power (**_hp_**) of the car and this 
relationship is approximated by the following linear equation/model:  
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_**

## Analysis  
Exploratory data analysis involved making box plots (Appendix) in which the 
efficiency (in **_mpg_**) was grouped by the transmission type
(**_am_**). From the box plots itself, 
it can be seen that manual transmission cars (*MT*/**_am_** = 1) are more 
efficient 
and give higher **_mpg_** than automatic transmission (*AT*/**_am_** = 0) 
cars. However, there is considerable variation in the **_mpg_** within each 
transmission type (**_am_**) and effects of other factors should also be 
analysed.
To begin with, the simplest of models is made, and this calculates as to how 
**_mpg_** varies by **_am_** (*AT/MT*). Computation 1 (Appendix) 
provides the first linear model: **_mpg=17.147+7.245(am)_**. The coefficients of 
this simple model show that a _MT_ car 
runs 24.392 **_mpg_** compared to 17.145 **_mpg_** for an _AT_ car.  

Any model that calculates car **_mpg_**, should also take into consideration, 
that **_mpg_** of any car is related to it's weight. *Should weight be included* 
*in the model?*   
In computation 2 (Appendix), a new model is made in which weight is included, 
residuals are tested for normalility and ANOVA done for the 
previous model and the new model.  
The null hypothesis (NH) proposed for ANOVA is that omitting **_wt_** will not 
increase the bias in the model. 
Alternative hypothesis (AH) is that omitting **_wt_** 
will increase the bias. 
From ANOVA, we find that the F score is 46.115 (P value <.0001), 
which is significant. We therefore, reject the 
NH, accept the AH and include **_wt_** to improve the model. 
The model changes to: **_mpg=37.322-0.024(am)-5.353(wt)_**  

Next we need to test, whether to include other engine variables 
(like horse power **_hp_**) that might effect **_mpg_**.  
In computation 3 (Appendix), **_hp_** is added to the model and evaluated. 
Including **_hp_** as a regressor in the model, increases variance of 
both the **_wt_** and **_am_** coefficients, thereby increasing the overall 
variance of the model. _Should_ **_hp_** _be included?_  
From the shapiro test, residuals of the updated model are normal so ANOVA can be 
performed. ANOVA tests the NH that omitting the **_hp_** will not increase 
bias in the model and gives us a F score of 15.224 (P value < .001), which is 
significant. The NH is rejected and **_hp_** is included in 
the model which now transforms to: 
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_**  

**_hp_** _normally depends upon_ **_disp_** _and_ **_cyl_**. _Should these_
_be included?_  
We include **_disp_** and **_cyl_** in the old model and evaluate it 
- computation 4 (Appendix). As suspected, since **_hp_** depends upon 
**_disp_** and **_cyl_**, it is strongly correlated with both and including them 
increases variances of the coefficients of **_am,wt,hp_**. 
From the shapiro test, residuals are normally distributed, ANOVA can be done. 
NH is the same as before i.e. omitting **_disp_** and **_cyl_** will not
increase 
the bias of the model. ANOVA of the two models gives F score of 1.369 
(P values .27) which is not significant. Also, there is no significant reduction 
in the RSS (residual sum of squares): 180 vs 163, between the old and new 
models.  
Therefore, we can accept the 
NH and omit **_disp_** and **_cyl_** from our model.  

Similarly, we include all variables (in addition to **_am , wt and hp_**) in the 
final testing model. Based on results of computation 5 (Appendix), 
it is inferred that
including all the other variables in the testing model, increases the variances
of the coefficients that matter but does not significantly 
reduce the RSS: 180 vs 147,  between the last model 
and the testing model. There is no gain in including any regressors
other than **_am , wt and hp_** in the model.  

## Conclusion  
Based on the analysis, the linear model specified by 
**_mpg = 34.003+2.084(am)-2.879(wt)-0.038(hp)_** with a standard error of 
2.538 **_mpg_** is the best fit to the data provided.  
Based on this model, we can 
conclude that a *MT* car will run 2.084 miles more than an *AT* car with the 
same **_wt/hp_**. Each ton increase in **_wt_** will decrease the car **_mpg_** by
2.879 **_mpg_** and 
one **_hp_** increase in engine power will decrease the car **_mpg_** by 0.038 mpg.   
Finally, the error between the **_mpg_** for any data point in the data set
and the **_mpg_** calculated using the model should not exceed +/- 2.538 **_mpg_**. 

## Appendix

**EDA-Box Plot**
```{r  chunk1}
data(mtcars)
require('car')
par(cex=0.7)
boxplot(mpg~factor(am),data=mtcars, main="Car Mileage Data",
    xlab="Auto (0/AT) or Manual (1/MT) Transmission ", ylab="Miles Per Gallon") 
```

**Computation 1**
```{r chunk2}
mdl1<-lm(mpg~factor(am),mtcars)
mdl1$coefficients
shapiro.test(mdl1$residuals)
```
**Computation 2**  
Shapiro tests the null hypothesis that the data provided is normal and if so 
ANOVA may be done to compare different models. 
```{r chunk3}
mdl2<-lm(mpg~factor(am)+wt,mtcars)
mdl2$coefficients
shapiro.test(mdl2$residuals)
anova(mdl1,mdl2)
```
**Computation 3** 
```{r chunk4}
mdl9<-lm(mpg~factor(am)+wt+hp,mtcars)
par(mfrow=c(1,2), cex=0.5)
plot(mdl9$residuals,main="Residual Plot",xlab="Sample No.", ylab="Residual" )
hist(mdl9$residuals, main="Histogram of Residuals",ylab="Frequency",
     xlab="Residual Value")
shapiro.test(mdl9$residuals)
```
  
Plot shows even dispersion of residuals and 
Histogram shows residuals are approximately normal. 
```{r chunk5}

vif(mdl2)
vif(mdl9)
anova(mdl2,mdl9)
summary(mdl9)
```
**Computation 4**
```{r chunk6}
mdl11<-lm(mpg~factor(am)+wt+hp+disp+cyl,mtcars)
vif(mdl11)
shapiro.test(mdl11$residuals)
anova(mdl9,mdl11)
```
**Computation 5**
```{r chunk7}
mdl12<-lm(mpg~factor(am)+wt+hp+disp+cyl+drat+qsec+factor(vs)+carb+gear,mtcars)
vif(mdl12)
shapiro.test(mdl12$residuals)
anova(mdl9,mdl12)
```









 
 